Online Multiclass Linear Classification with Bandit Feedback

### 2019-01-17 ICML-submission/introduction.tex
another pass over related work

\begin{protocol}[h]
\caption{\textsc{Online Multiclass Classification with Bandit Feedback}
\label{algorithm:game-protocol}}
\begin{algorithmic}[1]
{
\REQUIRE{Number of classes $K$, number of rounds $T$.}
\FOR{$t=1,2,\dots,T$}
\STATE Adversary chooses example $(x_t, y_t)$, where $x_t$ is revealed to the learner.
\STATE{Predict class label $\widehat y_t \in \{1,2,\dots,K\}$.}
\STATE{Observe feedback $z_t \in \{0,1\}$, \\ \qquad where $z_t = \begin{cases}
0 & \text{if $\widehat y_t = y_t$,} \\
1 & \text{if $\widehat y_t \neq y_t$.}
\end{cases}$
}
\ENDFOR
}
\end{algorithmic}
\end{protocol}

@@

In this paper, we focus on the special case when the examples chosen by the
adversary lie in $\R^d$ and are linearly separable with a margin. We introduce
two notions of linear separability, \emph{weak} and \emph{strong}, formally
stated in
Definitions~\ref{definition:weak-linear-separability}~and~\ref{definition:strong-linear-separability}.
The standard notion of multiclass linear separability corresponds to the weak
linear separability. For binary classification, the weak and strong separability
are identical. However, they differ for multiclass classification with $K \ge 3$
classes. For multiclass classification with $K$ classes, weak linear
separability means that the examples from each class lie in an intersection of
$K-1$ halfspaces and the examples outside of the class lie in the complement of
the intersection of the halfspaces. Strong linear separability means that
examples from each class are separated from the remaining examples by a
\emph{single} hyperplane.

In the full-information feedback setting, it is well known that if all examples
have norm at most $R$ and are weakly linearly separable with a margin $\gamma$,
then the \textsc{Multiclass Perceptron} algorithm makes at most $\lfloor
2(R/\gamma)^2 \rfloor$ mistakes. The guarantees provided by \textsc{Multiclass
Perceptron} is very satisfying, since it achieves an information-theoretically
optimal mistake upper bound, while having low time and memory
complexities.~\footnote{These results are standard in the literature. For
completeness, we present these results along with their proofs in
Appendix~\ref{section:multiclass-perceptron-proofs} in the supplementary
material.}

The bandit feedback setting, however, is much more challenging. For the case
when the examples are strongly linearly separable, to the best of our knowledge,
it is not known how to design an efficient algorithm with a finite mistake bound
before our work.~\footnote{Although~\cite{Chen-Chen-Zhang-Chen-Zhang-2009}
claimed that their Conservative OVA algorithm with PA-I update has a finite
mistake bound under the strong linear separability condition, their proof of
Theorem 2 is incorrect. Their Lemma 4 cannot be directly applied to the bandit
setting, as the algorithm only makes updates on a subset of the $w_i$'s.} We
design a simple and efficient algorithm, namely
Algorithm~\ref{algorithm:algorithm-for-strongly-linearly-separable-examples},
that has an expected number of mistakes at most $(K-1) \lfloor 4(R/\gamma)^2
\rfloor$. The memory complexity and per-round time complexity are both $O(dK)$.
The algorithm can be viewed as running $K$ copies of the \textsc{Binary
Perceptron} algorithm, one copy for each class. Intuitively speaking, the $O(K)$
multiplicative factor in the mistake bound is the price we pay for the bandit
feedback, or more precisely, the lack of full-information feedback. We prove
that any (possibly randomized) algorithm must make $\Omega(K (R/\gamma)^2))$
mistakes in the worst case.

For the case when examples are weakly linearly separable, we construct a
non-linear feature mapping $\phi$ and associated positive definite kernel
$k(x,x')$ that makes the examples \emph{strongly} linearly separable. We then
use the kernelized version of the algorithm for the strongly separable case,
Algorithm~\ref{algorithm:kernelized}. (For details on kernel methods see
e.g.~\citet{Scholkopf-Smola-2002} or \citet{Shawe-Taylor-Cristianini-2004}.)
The kernel $k(x,x')$ corresponding to the feature mapping $\phi$ has a simple
explicit formula and can be computed in $O(d)$ time.

@@

The resulting kernelized algorithm runs in time that is polynomial in the
original dimension of the feature vectors $d$, the number of classes $K$, and
the number of rounds $T$. We prove that if the examples lie in unit ball of
$\R^d$ and are weakly separable with margin $\gamma$, the algorithm makes at
most $2^{\widetilde{O}(\min(K \log^2 \frac{1}{\gamma}, \sqrt{\frac{1}{\gamma}}
\log K))}$ mistakes.

>>>>

\begin{protocol}[h]
\caption{\textsc{Online Multiclass Classification with Bandit Feedback \chicheng{Shall we emphasize linear here?}}
\label{algorithm:game-protocol}}
\begin{algorithmic}[1]
{
\REQUIRE{Number of classes $K$, number of rounds $T$.}
\FOR{$t=1,2,\dots,T$}
\STATE Adversary chooses example $(x_t, y_t)$, where $x_t$ is revealed to the learner.
\STATE{Predict class label $\widehat y_t \in \{1,2,\dots,K\}$.}
\STATE{Observe feedback $z_t \in \{0,1\}$, \\ \qquad where $z_t = \begin{cases}
0 & \text{if $\widehat y_t = y_t$,} \\
1 & \text{if $\widehat y_t \neq y_t$.}
\end{cases}$
}
\ENDFOR
}
\end{algorithmic}
\end{protocol}

@@

In this paper, we focus on the special case when the examples chosen by the
adversary lie in $\R^d$ and are linearly separable with a margin. We introduce
two notions of linear separability, \emph{weak} and \emph{strong}, formally
stated in
Definitions~\ref{definition:weak-linear-separability}~and~\ref{definition:strong-linear-separability}.
The standard notion of multiclass linear separability~\citep[See e.g.][]{Crammer-Singer-2003}
corresponds to the weak
linear separability. For binary classification, the weak and strong separability
are identical. However, they differ for multiclass classification with $K \ge 3$
classes. For multiclass classification with $K$ classes, weak linear
separability means that the examples from each class lie in an intersection of
$K-1$ halfspaces and the examples outside of the class lie in the complement of
the intersection of the halfspaces. Strong linear separability means that
examples from each class are separated from the remaining examples by a
\emph{single} hyperplane.

In the full-information feedback setting, it is well known that if all examples
have norm at most $R$ and are weakly linearly separable with a margin $\gamma$,
then the \textsc{Multiclass Perceptron} algorithm makes at most $\lfloor
2(R/\gamma)^2 \rfloor$ mistakes.
\chicheng{I just noted that Theorem 3 of~\cite{Crammer-Singer-2003} actually
has exactly the same upper bound - but it might still make sense to include the
proof here.}
The guarantees provided by \textsc{Multiclass
Perceptron} is very satisfying, since it achieves an information-theoretically
optimal mistake upper bound, while having low time and memory
complexities.~\footnote{These results are standard in the literature. For
completeness, we present these results along with their proofs in
Appendix~\ref{section:multiclass-perceptron-proofs} in the supplementary
material.}

The bandit feedback setting, however, is much more challenging. For the case
when the examples are strongly linearly separable, to the best of our knowledge,
it is not known how to design an efficient algorithm with a finite mistake bound
before our work.~\footnote{Although~\cite{Chen-Chen-Zhang-Chen-Zhang-2009}
claimed that their Conservative OVA algorithm with PA-I update has a finite
mistake bound under the strong linear separability condition, their proof of
Theorem 2 is incorrect. Their Lemma 4 cannot be directly applied to the bandit
setting, as the algorithm only makes updates on a subset of the $w_i$'s.} We
design a simple and efficient algorithm, namely
Algorithm~\ref{algorithm:algorithm-for-strongly-linearly-separable-examples},
that has an expected number of mistakes at most $O(K (R/\gamma)^2))$.
The memory complexity and per-round time complexity are both $O(dK)$.
The algorithm can be viewed as running $K$ copies of the \textsc{Binary
Perceptron} algorithm, one copy for each class. Intuitively speaking, the $O(K)$
multiplicative factor in the mistake bound is the price we pay for the bandit
feedback, or more precisely, the lack of full-information feedback. We prove
that any (possibly randomized) algorithm must make $\Omega(K (R/\gamma)^2))$
mistakes in the worst case.
%$(K-1) \lfloor 4(R/\gamma)^2\rfloor$

For the case when the examples are weakly linearly separable, we construct a
non-linear feature mapping $\phi$ and associated positive definite kernel
$k(x,x')$ that makes the examples \emph{strongly} linearly separable. We then
use the kernelized version of
Algorithm~\ref{algorithm:algorithm-for-strongly-linearly-separable-examples}
for the strongly separable case, namely
Algorithm~\ref{algorithm:kernelized}. (For details on kernel methods see
e.g.~\citet{Scholkopf-Smola-2002} or \citet{Shawe-Taylor-Cristianini-2004}.)
The kernel $k(x,x')$ corresponding to the feature mapping $\phi$ has a simple
explicit formula and can be computed in $O(d)$ time,
making Algorithm~\ref{algorithm:kernelized} computationally efficient.

@@

The resulting kernelized algorithm runs in time polynomial in the
original dimension of the feature vectors $d$, the number of classes $K$, and
the number of rounds $T$. We prove that if the examples lie in the unit ball of
$\R^d$ and are weakly linearly separable with margin $\gamma$,
Algorithm~\ref{algorithm:kernelized} makes at
most $2^{\widetilde{O}(\min(K \log^2 \frac{1}{\gamma}, \sqrt{\frac{1}{\gamma}}
\log K))}$ mistakes.

### 2019-01-17 ICML-submission/related-work.tex
another pass over related work

The problem of online bandit multiclass learning was initially formulated in the
pioneering work of~\citet{Auer-Long-1999}, under the name of ``weak
reinforcement model''. They showed that if all examples agree with some
classifier from a prespecified hypothesis class $\calH$, then the optimal
mistake bound in the bandit setting can be upper bounded by the optimal mistake
bound in the full information setting, times a factor of $(2.01 + o(1))K \ln K$.
\citet{Long-2017} later improved the factor to $(1 + o(1)) K \ln K$ and showed
its near-optimality. \citet{Daniely-Helbertal-2013} extended the results to the
setting where the performance of the algorithm is measured by its regret, i.e.
the difference between the number of mistakes made by the algorithm and the
number of mistakes made by the best classifier in $\calH$ in hindsight. However,
all algorithms developed in this context are computationally inefficient.

The linear classification version of this problem is initially studied
by~\citet{Kakade-Shalev-Shwartz-Tewari-2008}. They proposed two computationally
inefficient algorithms that work in the weakly linearly separable setting, one
with a mistake bound of $O(K^2 d \ln(\frac{d}{\gamma}))$, the other with a
mistake bound of $\widetilde{O}(\frac{K^2}{\gamma^2} \ln T)$. The latter result
was later improved by \citet{Daniely-Helbertal-2013}, which gives a
computationally inefficient algorithm with a mistake upper bound of
$\widetilde{O}(\frac{K}{\gamma^2})$. In addition, the authors propose the
\textsc{Banditron} algorithm, a computationally efficient algorithm that has a
$O(T^{2/3})$ regret against the multiclass hinge loss in the general setting,
and has a $O(\sqrt{T})$ mistake bound in the $\gamma$-weak linearly separable
setting. The polynomial dependence of \textsc{Banditron}'s mistake bound on the
time horizon is undesirable for problems with a long time horizon. One key open
question left by~\citet{Kakade-Shalev-Shwartz-Tewari-2008} is whether one can
design computationally efficient algorithms that achieve mistake bounds that
match or improve over those of inefficient algorithms. In this paper, we take a
step towards answering this question, showing that efficient algorithms with
mistake bounds quasipolynomial in $\frac 1 \gamma$ or quasipolynomial in $K$ can
be obtained.

@@

Many works consider bandit multiclass classification in the general setting,
where the data is not necessarily separable by some linear
classifier~\cite{Abernethy-Rakhlin-2009, Wang-Jin-Valizadegan-2010,
Crammer-Gentile-2013, Hazan-Kale-2011, Beygelzimer-Orabona-Zhang-2017,
Foster-Kale-Luo-Mohri-Sridharan-2018}. In addition, \citet{Chen-Lin-Lu-2014,
Zhang-Jung-Tewari-2018} study online bandit multiclass boosting under bandit
feedback, where one can view boosting as linear classification by treating each
base hypothesis as a separate feature. However, most of these works achieve a
regret bound of order $\widetilde{O}(\sqrt{T})$ to $O(T^{3/4})$. In the weakly
separable setting, these algorithms only guarantees a mistake upper bound of
$\widetilde{O}(\sqrt{T})$, which is undesirable for problems with a long time
horizon. One exception is the \textsc{Newtron} algorithm, where with appropriate
settings of parameter $\alpha$, it is able to achieve a $O(\ln T)$ regret.
However, as argued in ~\citep[][Appendix D]{Beygelzimer-Orabona-Zhang-2017}, if
the setting of $\alpha$ makes the regret bound $O(\sqrt{T})$, the logistic loss
of the best linear predictor would be $\Omega(T^{3/4})$. Therefore,
\textsc{Newtron} is still only able to guarantee a mistake upper bound of
$O(\sqrt{T})$ in the weakly separable setting.

@@

There is a large body of literature on the contextual bandit learning
problem~\citep{Auer-2003, Langford-Zhang-2008}. In this problem, the learner is
given a policy class $\Pi$, and at each timestep $t$, an example $x_t$ is shown,
with an associated cost vector $c_t$ hidden. The learner then takes an action
$\widehat{y}_t$, and observes the incurred cost $c_t(\widehat{y}_t)$. The goal
of the learner is to minimize its regret $\sum_{t=1}^T c_t(\widehat{y}_t) -
\min_{\pi \in \Pi} \sum_{t=1}^T c_t(\pi(x_t))$. The online bandit multiclass
learning problem is a special case of the contextual bandits problem, where the
cost $c_t(i)$ is the classification error $\indicator{i \neq y_t}$, and the
policy class $\Pi$ is the set of linear classifiers $\cbr{x \mapsto \argmax_y
(Wx)_y: W \in \R^{k \times d}}$. A series of works in the contextual bandits
learning literature focus on
oracle-efficiency~\citep{Dudik-Hsu-Kale-Karampatziakis-Langford-Reyzin-Zhang-2011,
Agarwal-Hsu-Kale-Langford-Li-Schapire-2014, Rakhlin-Sridharan-2016,
Syrgkanis-Krishnamurthy-Schapire-2016,
Syrgkanis-Luo-Krishnamurthy-Schapire-2016}. Specifically, they assume access to
a policy optimization oracle that can return the policy in $\Pi$ that minimizes
its empirical cost on any set of cost-sensitive examples, and the algorithm only
calls the oracle a polynomial number of times. However, these algorithms are not
truly computationally efficient in our setting, as it is NP-hard in general to
find a linear classifier that minimizes the empirical cost over a set of
examples~\citep{Arora-Babai-Stern-Sweedyk-1997}.

@@

Recently, \citet{Foster-Krishnamurthy-2018} developed a rich theory of
contextual bandits with surrogate losses, focusing on regrets of the form
$\sum_{t=1}^T c_t(\widehat{y}_t) - \min_{f \in \calF} \sum_{t=1}^T \frac{1}{K}
\sum_{i=1}^K c_t(i) \phi( f_i(x_t) )$, where $\calF$ contains score functions
$f$ such that $\sum_{i=1}^K f_i(\cdot) \equiv 0$, and $\phi(s) = \max(1 - \frac
s \gamma, 0)$ or $\min(1, \max(1 - \frac s \gamma, 0))$. On one hand, it gives
information-theoretic mistake upper bounds under various settings of $\calF$,
including parametric and nonparametric classes. On the other hand, it gives an
efficient algorithm that has a $O(\sqrt{T})$ regret against the benchmark of
$\calF = \cbr{x \mapsto W x: W \in \R^{k \times d}, \one^T W = 0}$. A direct
application of this result to \textsc{Online Bandit Multiclass Linear
Classification} yields an algorithm with a $O(\sqrt{T})$ mistake bound in the
strongly linearly separable setting.

>>>>

The problem of online bandit multiclass learning was initially formulated in the
pioneering work of~\citet{Auer-Long-1999}, under the name of ``weak
reinforcement model''. They showed that if all examples agree with some
classifier from a prespecified hypothesis class $\calH$, then the optimal
mistake bound in the bandit setting can be upper bounded by the optimal mistake
bound in the full information setting, times a factor of $(2.01 + o(1))K \ln K$.
\citet{Long-2017} later improved the factor to $(1 + o(1)) K \ln K$ and showed
its near-optimality. \citet{Daniely-Helbertal-2013} extended the results to the
setting where the performance of the algorithm is measured by its regret, i.e.
the difference between the number of mistakes made by the algorithm and the
number of mistakes made by the best classifier in $\calH$ in hindsight.
We remark that
all algorithms developed in this context are computationally inefficient.

The linear classification version of this problem is initially studied
by~\citet{Kakade-Shalev-Shwartz-Tewari-2008}. They proposed two computationally
inefficient algorithms that work in the weakly linearly separable setting, one
with a mistake bound of $O(K^2 d \ln(\frac{d}{\gamma}))$, the other with a
mistake bound of $\widetilde{O}(\frac{K^2}{\gamma^2} \ln T)$. The latter result
was later improved by \citet{Daniely-Helbertal-2013}, which gives a
computationally inefficient algorithm with a mistake upper bound of
$\widetilde{O}(\frac{K}{\gamma^2})$. In addition, the authors propose the
\textsc{Banditron} algorithm, a computationally efficient algorithm that has a
$O(T^{2/3})$ regret against the multiclass hinge loss in the general setting,
and has a $O(\sqrt{T})$ mistake bound in the $\gamma$-weakly linearly separable
setting.
In contrast to mild dependencies on the time horizon for mistake bounds of
computationally inefficient algorithms,
the polynomial dependence of \textsc{Banditron}'s mistake bound on the
time horizon is undesirable for problems with a long time horizon, in the
weakly linearly separable setting. One key open
question left by~\citet{Kakade-Shalev-Shwartz-Tewari-2008} is whether one can
design computationally efficient algorithms that achieve mistake bounds that
match or improve over those of inefficient algorithms. In this paper, we take a
step towards answering this question, showing that efficient algorithms with
mistake bounds quasipolynomial in $\frac 1 \gamma$ or quasipolynomial in $K$ can
be obtained.

@@

Many works consider bandit multiclass classification in the general setting,
where the examples are not necessarily separable by some linear
classifier~\cite{Abernethy-Rakhlin-2009, Wang-Jin-Valizadegan-2010,
Crammer-Gentile-2013, Hazan-Kale-2011, Beygelzimer-Orabona-Zhang-2017,
Foster-Kale-Luo-Mohri-Sridharan-2018}. In addition, \citet{Chen-Lin-Lu-2014,
Zhang-Jung-Tewari-2018} study online bandit multiclass boosting under bandit
feedback, where one can view boosting as linear classification by treating each
base hypothesis as a separate feature. However, most of these works achieve a
regret of order $\widetilde{O}(\sqrt{T})$ to $O(T^{3/4})$.
In the weakly linearly
separable setting, these algorithms at best guarantees a mistake upper bound of
$\widetilde{O}(\sqrt{T})$, which is undesirable for problems with a long time
horizon. One exception is the \textsc{Newtron} algorithm, where with appropriate
settings of parameter $\alpha$, it is able to achieve a $O(\ln T)$ regret against
the $\alpha$-logistic loss.
However, as argued in ~\citep[][Appendix D]{Beygelzimer-Orabona-Zhang-2017}, if
the setting of $\alpha$ makes the regret bound $O(\sqrt{T})$, the $\alpha$-logistic loss
of the best linear predictor would be $\Omega(T^{3/4})$. Therefore,
\textsc{Newtron} at best guarantees a mistake upper bound of
$O(\sqrt{T})$ in the weakly linearly separable setting.

@@

There is a large body of literature on the contextual bandits
problem~\citep{Auer-2003, Langford-Zhang-2008}. In this problem, the learner is
given a policy class $\Pi$, and at each timestep $t$, an example $x_t$ is shown,
with an associated cost vector $c_t$ hidden. The learner then takes an action
$\widehat{y}_t$, and observes the incurred cost $c_t(\widehat{y}_t)$. The goal
of the learner is to minimize its regret $\sum_{t=1}^T c_t(\widehat{y}_t) -
\min_{\pi \in \Pi} \sum_{t=1}^T c_t(\pi(x_t))$. The online bandit multiclass
learning problem can be seen a special case of the contextual bandits problem, where the
cost $c_t(i)$ is the classification error $\indicator{i \neq y_t}$, and the
policy class $\Pi$ is the set of linear classifiers $\cbr{x \mapsto \argmax_y
(Wx)_y: W \in \R^{k \times d}}$. A series of works in the contextual bandits
learning literature focus on
oracle-efficiency~\citep{Dudik-Hsu-Kale-Karampatziakis-Langford-Reyzin-Zhang-2011,
Agarwal-Hsu-Kale-Langford-Li-Schapire-2014, Rakhlin-Sridharan-2016,
Syrgkanis-Krishnamurthy-Schapire-2016,
Syrgkanis-Luo-Krishnamurthy-Schapire-2016}. Specifically, they assume access to
a policy optimization oracle that can return the policy in $\Pi$ that minimizes
its empirical cost on any set of cost-sensitive examples, and the algorithm only
calls the oracle a polynomial number of times. However, these algorithms are not
truly computationally efficient in our setting, as it is NP-hard in general to
find a linear classifier that minimizes the empirical cost over a set of
examples~\citep{Arora-Babai-Stern-Sweedyk-1997}.

@@

Recently, \citet{Foster-Krishnamurthy-2018} developed a rich theory of
contextual bandits with surrogate losses, focusing on regrets of the form
$\sum_{t=1}^T c_t(\widehat{y}_t) - \min_{f \in \calF} \sum_{t=1}^T \frac{1}{K}
\sum_{i=1}^K c_t(i) \phi( f_i(x_t) )$, where $\calF$ contains score functions
$f$ such that $\sum_{i=1}^K f_i(\cdot) \equiv 0$, and $\phi(s) = \max(1 - \frac
s \gamma, 0)$ or $\min(1, \max(1 - \frac s \gamma, 0))$. On one hand, it gives
information-theoretic regret upper bounds under various settings of $\calF$.
On the other hand, it gives an
efficient algorithm that has a $O(\sqrt{T})$ regret against the benchmark of
$\calF = \cbr{x \mapsto W x: W \in \R^{k \times d}, \one^T W = 0}$. A direct
application of this result to \textsc{Online Bandit Multiclass Linear
Classification} yields an algorithm with a $O(\sqrt{T})$ mistake bound in the
strongly linearly separable setting.

%including parametric and nonparametric classes
